Tutorial_2a

### 2024-11-23 Tutorial_2a.Rmd
update tut 2a + add png

@@

@@

```{r priors-nonlinearsyntax}
priors <- c(
  prior(skew_normal(-0.59,1.26,-4.22), class = b, nlpar = a), 
  prior(skew_normal(-0.2,.71,-2.2), class = b, nlpar = b), 
  prior(skew_normal(-0.2,.71,-2.2), class = b, nlpar = c), 
  prior(exponential(1), class = sd, group = pid, nlpar = a),
  prior(exponential(1), class = sd, group = pid, nlpar = b),
  prior(lkj(2), class = cor, group = pid)
)
```

@@

For the average participant, using add_epred_draws().

@@

Plot of posterior distributions of sigma(timebinX:primeY) : check size differences between timebins and prime types?

@@

Future studies could (a) increase the number of participants to estimate the porportion of "dippers" in the subject population, and/or (b) try to explain why this dip occurs. For example, Panis and Schmidt (ref) concluded that active, top-down, task-guided response inhibition effects emerge around 360 ms after the onset of the stimulus following the prime (here:the target). Such a top-down inhibitory effect might exist in our priming data set, because after some time participants will learn that the first stimulus is not the one they have to respond to; To prevent a premature overt response to the prime they thus might gradually increase a global response threshold during the remainder of the experiment, which could result in a lower hazard in congruent trials compared to blank trials, for bins after ~360 ms, and towards the end of the experiment.
Interestingly, all subjects show a tendency in their mean difference (congruent minus blank) to "dip" around that time. Therefore, future modeling efforts could incorporate the trial number into the model formula, in order to also study how the effects of prime type on hazard change on the experiment-wide time scale, next to the trial-wide time scale (TIME). In Tutorial_2a.Rmd we provide a number of model formula that should get you going.

>>>>

Perform a prior predictive check.

```{r}
plan(multicore)

model_M0_prior <-                     # wrong data file name
   brm(data = M0_data,
       family = bernoulli(link="cloglog"),
       event ~ 0 + timebin + (0 + timebin | pid),
       prior = priors_M0,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = 'only',
       file = "Tutorial_2_Bayesian/models/model_M0_prior")
```

```{r}

#pp_check(model_M0_prior, stat="mean")

#http://bruno.nicenboim.me/bayescogsci/

# prior predictive check for 1 time bin
cloglog_predictive_distribution <-
  function(samples, N_obs) {
    # empty data frame with headers:
    df_pred <- tibble(trial = numeric(0),
                      y_pred = numeric(0),
                      iter = numeric(0))
    # i iterates from 1 to the length of samples,
      for (i in seq_along(samples)) {
        cloglog_haz <- samples[i]
        df_pred <- bind_rows(df_pred,
                             tibble(trial = 1:N_obs,
                                    y_pred = rbernoulli(N_obs, 1-exp((-1)*exp(cloglog_haz))),
                                    iter = i))
    }
    df_pred # return
  }
# simulate prior hazard values for 1 bin
N_samples <- 3000
N_obs <- 100 #nrow(M0_data)
#cloglog_samples <- rskew_normal(N_samples,-0.59,1.26,-4.22) # flat prior for cloglog intercept
cloglog_samples <- rskew_normal(N_samples,-1,1,-2) # weakly informative prior for cloglog intercept

prior_pred <- cloglog_predictive_distribution(cloglog_samples,N_obs)

prior_pred_haz <- prior_pred %>% group_by(iter) %>% summarise(pred_haz = mean(y_pred))

# plot 
ggplot(prior_pred_haz, aes(x=pred_haz)) +
  geom_histogram(binwidth=.01) +
  scale_x_continuous(limits = c(0,1))
```

```{r}
# Generate prior predictive samples for 10 bins 
bins_cloglog_predictive_distribution <-
  function(samples_bin, N_obs, N_bins) {
    # empty data frame with headers:
    df_pred <- tibble(trial = numeric(0),
                      y_pred = numeric(0),
                      iter = numeric(0),
                      bin = numeric(0))
    # i iterates from 1 to the length of samples,
    # b interates from 1 to N_bins
    for( b in 1:N_bins){
      for (i in 1:dim(samples_bin)[2]) {
        cloglog_haz <- samples_bin[b,i]
        df_pred <- bind_rows(df_pred,
                             tibble(trial = 1:N_obs,
                                    y_pred = rbernoulli(N_obs, 1-exp((-1)*exp(cloglog_haz))),
                                    iter = i,
                                    bin = b))
      }
    }
    df_pred # return
  }

# simulate prior hazard functions for 10 bin
N_samples <- 40
N_obs <- 100
N_bins <- 10
#cloglog_samples <- rskew_normal(N_samples,-0.59,1.26,-4.22) # flat prior for cloglog intercept
cloglog_samples_bin <- data.frame()
for(b in 1:N_bins){
  cloglog_samples_bin[b,1:N_samples] <- rskew_normal(N_samples,-1,1,-2) # weakly informative prior for cloglog intercept
}

prior_pred <- bins_cloglog_predictive_distribution(cloglog_samples_bin,N_obs,N_bins)

prior_pred_haz <- prior_pred %>% group_by(iter,bin) %>% summarise(pred_haz = mean(y_pred))

# plot 
ggplot(prior_pred_haz, aes(x=bin, y=pred_haz, group=iter)) +
  #geom_histogram(binwidth=.01) +
  geom_line(color="black") +
  geom_line(data = prior_pred_haz %>% filter(iter==4), color="red", linewidth = 2) +
  geom_line(data = prior_pred_haz %>% filter(iter==18), color="blue", linewidth = 2) +

  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(1,10), breaks = c(1:10))

```

```{r}
# Generate prior predictive hazard functions for 6 participants
newdata_prior = tibble(timebin = c(6:15)) %>% 
                            expand_grid(pid = 1:6)
N_obs = 100

# prepare columns of data set
df_pred <- tibble(timebin = integer(0),
                  pid = integer(0),
                  .row = integer(0),
                  .chain = integer(0),
                  .iteration = integer(0),
                  .draw = integer(0),
                  .prediction = integer(0),
                  obs = integer(0))

for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0_prior, newdata = newdata_prior, summary=F, ndraws=NULL)
  prior_pred <- prior_pred %>% mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}
head(df_pred)

# calculate hazard per draw (average across obs)
df_pred_haz <- df_pred %>% group_by(pid,timebin,.draw) %>% summarise(pred_haz = mean(.prediction))
summary(df_pred_haz)

# plot 40 draws per participant
ggplot(df_pred_haz %>% filter(.draw < 21), aes(x=timebin, y=pred_haz, group=.draw)) +

   geom_line(color="black") +
   geom_line(data=df_pred_haz %>% filter(.draw == 20), color="red", linewidth=2) +

  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(6,15), breaks = c(6:15)) +
  facet_wrap(~pid)

```

@@

## Intermezzo: Prior predictive checks

model_M0c_prior <-
   brm(data = M0c_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + period_9 + period_9_sq + period_9_cu + 
                          (0 + Intercept + period_9 + period_9_sq + period_9_cu  | pid),
       prior = priors_M0c,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior")
```

```{r}
# Generate prior predictive hazard functions for 6 participants
newdata_prior = tibble(period_9 = c(-3:6)) %>% 
  mutate(period_9_sq = period_9^2,
         period_9_cu = period_9^3) %>%
  expand_grid(pid = 1:6)

N_obs = 100

# prepare columns of data set
df_pred <- tibble(period_9 = integer(0),
                  period_9_sq = double(0),
                  period_9_cu = double(0),
                  pid = integer(0),
                  .row = integer(0),
                  .chain = integer(0),
                  .iteration = integer(0),
                  .draw = integer(0),
                  .prediction = integer(0),
                  obs = integer(0))

for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0c_prior, newdata = newdata_prior, summary=F, ndraws=NULL)
  prior_pred <- prior_pred %>% mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}
head(df_pred)

# calculate hazard per draw (average across obs)
df_pred_haz <- df_pred %>% group_by(pid,period_9,.draw) %>% summarise(pred_haz = mean(.prediction))
summary(df_pred_haz)

# plot 40 draws per participant
ggplot(df_pred_haz %>% filter(.draw < 2), aes(x=period_9, y=pred_haz, group=.draw)) +

   geom_line(color="black") +
   geom_line(data=df_pred_haz %>% filter(.draw == 17), color="red", linewidth=2) +

  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```

##

```{r}
dat <- tibble(x = -3:6,
              x2 = x^2,
              x3 = x^3,
              cloglog1 = -1.8   + 0.3*x   - 0.1*x2   + 0.009*x3,
              cloglog2 = -1     + 0.5*x   - 0.1*x2   + 0.004*x3,
              cloglog3 = -3     - 0.1*x   + 0.12*x2  - 0.002*x3,
              cloglog4 = -0.40  + 0.2*x   - 0.04*x2  + 0.003*x3,
              haz1 = 1 - exp(-1*exp(cloglog1)),
              haz2 = 1 - exp(-1*exp(cloglog2)),
              haz3 = 1 - exp(-1*exp(cloglog3)),
              haz4 = 1 - exp(-1*exp(cloglog4)))

p1<-ggplot(dat, aes(x=x)) +  
      geom_line(aes(y=cloglog1), color="black") +
      geom_line(aes(y=cloglog2), color="green") +
      geom_line(aes(y=cloglog3), color="red") +
      geom_line(aes(y=cloglog4), color="blue") +
      scale_y_continuous(limits=c(-6,2)) +
      scale_x_continuous(breaks=c(-3:6))

p2<-ggplot(dat, aes(x=x))+
      geom_line(aes(y=haz1), color="black") + 
      geom_line(aes(y=haz2), color="green") +
      geom_line(aes(y=haz3), color="red") +
      geom_line(aes(y=haz4), color="blue") +
      scale_y_continuous(limits=c(0,1)) +
      scale_x_continuous(breaks=c(-3:6))

p1|p2
```

```{r}
tibble(cloglog = rnorm(1000, .1, .1)) %>%
  ggplot(aes(x=cloglog)) +
  geom_density()
```

```{r data-priors-M0c}
M0c_data <- ptb_data %>% select(pid, event, period_9) %>%
  mutate(period_9_sq = period_9^2)

priors_M0c_prior <- c(
  #set_prior("skew_normal(-0.2,0.71,-2.2)", class = "b"),      
 set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),

#set_prior("normal(0.5,.5)",class="b",coef="period_9"),
#set_prior("normal(-.1,.2)",class="b",coef="period_9_sq"),   # Np3

set_prior("normal(0.1,.08)",class="b",coef="period_9"), # ok !!!
set_prior("normal(0,.06)",class="b",coef="period_9_sq"),

#set_prior("skew_normal(0.3,0.1,2)",class="b",coef="period_9"), #skNp3
#set_prior("skew_normal(0,0.1,0)",class="b",coef="period_9_sq"),

 set_prior("normal(0, .01)", class = "sd"),  
  set_prior("lkj(2)", class = "cor")  
)
```

model_M0_lin_prior448 <-
   brm(data = M0c_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + period_9 + period_9_sq +
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_prior,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0_lin_prior448")
```

```{r}
post <- as_draws_df(model_M0_lin_prior448) %>% 
  select(-lp__) %>% 
  as_tibble()

tidy_fixed <- post %>% 
  select(starts_with("b_"), .chain, .iteration, .draw) %>% 
  rename(chain=.chain, iter=.iteration, draw=.draw) %>% 
  pivot_longer(-c(chain, draw, iter)) 
head(tidy_fixed)
tail(tidy_fixed)

# plot
p_tidy_fixed <- ggplot(tidy_fixed, aes(x = value, y = name)) +  
  stat_halfeye(alpha=0.7,point_interval = "median_qi",.width=c(.8,.99) ) 
p_tidy_fixed

```

```{r}
#model_M0_lin_prior14 <-readRDS("Tutorial_2_Bayesian/models/model_M0c_lin_prior14.rds")               
# Generate prior predictive hazard functions for 6 participants
newdata_prior = tibble(period_9 = c(-3:6)) %>% 
  mutate(period_9_sq = period_9^2) %>%
  expand_grid(pid = 1:6)

N_obs = 150 # increase for smoother results

# prepare columns of data set
df_pred <- tibble(period_9 = integer(0),
                 period_9_sq = integer(0),
                 pid = integer(0),
                  .row = integer(0),
                  .chain = integer(0),
                  .iteration = integer(0),
                  .draw = integer(0),
                  .prediction = integer(0),
                  obs = integer(0))

set.seed(1)

for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0_lin_prior448, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=NULL)
  prior_pred <- prior_pred %>% 
    mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

head(df_pred)

# calculate hazard per draw (average across obs)
df_pred_haz <- df_pred %>% 
  group_by(pid,period_9,.draw) %>% 
  summarise(pred_haz = mean(.prediction)) 
summary(df_pred_haz)

# plot 40 draws per participant
sel = sample(1:8000,3,replace=F)

ggplot(df_pred_haz %>% 
   filter(.draw <7639), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   geom_line(data=df_pred_haz %>% filter(.draw ==7418), 
             color="red", linewidth=2) +
  geom_line(data=df_pred_haz %>% filter(.draw == 6001), 
            color="green", linewidth=2) +
  geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```

@@

```{r priors-nonlinearsyntax}
priors <- c(
  prior(skew_normal(-0.59,1.26,-4.22), class = b, nlpar = a), 
  prior(skew_normal(-0.2,.71,-2.2), class = b, nlpar = b), 
  prior(skew_normal(-0.2,.71,-2.2), class = b, nlpar = c), 
  prior(exponential(1), class = sd, group = pid, nlpar = a),
  prior(exponential(1), class = sd, group = pid, nlpar = b),
  prior(exponential(1), class = sd, group = pid, nlpar = c),
  prior(lkj(2), class = cor, group = pid)
)
```

@@

First, for the average participant, using add_epred_draws().

@@

[[todo or not todo]]
Plot of posterior distributions of sigma(timebinX:primeY) : check size differences between timebins and prime types? -> these are also population-level effects

@@

Future studies could (a) increase the number of participants to estimate the porportion of "dippers" in the subject population, and/or (b) try to explain why this dip occurs. For example, Panis and Schmidt (ref) concluded that active, top-down, task-guided response inhibition effects emerge around 360 ms after the onset of the stimulus following the prime (here:the target). Such a top-down inhibitory effect might exist in our priming data set, because after some time participants will learn that the first stimulus is not the one they have to respond to; To prevent a premature overt response to the prime they thus might gradually increase a global response threshold during the remainder of the experiment, which could result in a lower hazard in congruent trials compared to blank trials, for bins after ~360 ms, and towards the end of the experiment.
Interestingly, all subjects show a tendency in their mean difference (congruent minus blank) to "dip" around that time. Therefore, future modeling efforts could incorporate the trial number into the model formula, in order to also study how the effects of prime type on hazard change on the long experiment-wide time scale, next to the short trial-wide time scale (TIME). In Tutorial_2a.Rmd we provide a number of model formula that should get you going.
